Self-assembly of colloid-cholesteric composites: A route to switchable optical materials

### 2013-10-03 papers/2012-sandwich/CholColloids.tex
Changed references and electric field part and captionCholColloids.tex
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@2170 40876441-3496-0410-a613-a2fca35b2ae2

When spherical colloidal particles are mixed into a nematic liquid crystal, 
they disrupt the orientational order in the fluid and create defects (disclination lines) in the nematic close to their surfaces. To minimize the free energy cost, it is generally advantageous for defects to be shared:
therefore the colloidal particles have a generic tendency to aggregate. Such aggregates may further 
self-assemble into lines~\cite{wiresmiha}, 2D crystals~\cite{zumer}, 
planar structures~\cite{tanaka}, or 3D amorphous glasses~\cite{tiffany}.
Besides being of fundamental interest to materials science, these
structures have tunable elastic and optical properties~\cite{stark}. Hence 
they offer exciting prospects for applications as biosensors~\cite{abbott}, or
as new-generation display devices as they can exploit the bistability
inherent in liquid crystal devices~\cite{colloiddevice,choldevice}.

But one may also choose to disperse colloidal particles in a cholesteric, 
rather than nematic, liquid crystal. The molecules making up this material are
chiral, and this causes their average orientation (described by a director field) to rotate in space in
a helical fashion, rather than remain uniform as in nematics. 
This typically creates a 1-dimensional periodic structure, called the cholesteric phase, whose wavelength 
is the pitch, $p$. The periodicity can however become fully three dimensional in the so-called blue phases (BPs)~\cite{mermin}. BPs arise because the director field can twist around more than one direction at once (see 
the cartoon in Fig. S1). Such double twist 
regions (generally cylinders: DTCs) are energetically favoured at high enough molecular chirality, but
lead to geometric frustration because it is impossible to tile the whole 3D
space with DTCs without also introducing disclination lines: singular topological defects on which the director field is undefined. In blue phases these disclination lines themselves either form a 3D periodic regular lattice (in so called BPI and BPII), or remain disordered (BPIII)~\cite{bp3}. 
%
BPs are stabilised by the tendency of chiral molecules to twist, but destabilised by the cost of forming topological defects (which locally diminish the molecular alignment). In conventional cholesteric materials~\cite{mermin} this balance is only achievable in a narrow temperature range of a few degrees, around the onset of liquid crystalline ordering. However, recent advances in formulation have widened this stability range enormously (up to 250K) paving the way to
applications of BPs in display devices~\cite{kikuchi,coleswidetrange,bpdevice}.

@@

A second compelling reason to study colloid-BP composites is that, even without particles, BPs can show several competing metastable free energy minima,
corresponding to different topologies of the defect network~\cite{adriano}, with a strong dependence on applied electric or magnetic fields~\cite{henrichfield}.
% 
Adding particles is likely to further enrich the free energy landscape, creating composites that might be promising candidates for bistable or multistable devices, in which energy is needed only to switch optical properties and not to maintain them. (This is the e-paper paradigm~\cite{epaper}, and can lead to huge energy savings.)
To explore such metastability, and switching strategies between states, we require multi-unit-cell, time-dependent simulations. These go beyond previous studies of colloid-BP composites~\cite{miha} but follow comparable simulations of pure BPs~\cite{domaingrowth,bp3,henrichfield}, and isolated and dimeric colloids in cholesterics~\cite{juho1,juho2}. Our methodology and simulation parameters are summarized in the Supporting Information.

We show in Fig.~1 A-D the final state structures of four simulations in which we first equilibrated a stable BPI disclination network in bulk (with periodic boundary conditions) and then dispersed colloidal particles within it. Among dimensionless control parameters are the colloid volume fraction $\phi$  ($\phi = 1\%$ or $5\%$); the ratio $r = R/\lambda$ of particle radius to the BP lattice parameter $\lambda$ (here $r\simeq 0.10$) and the ratio $w = WR/K$ where $W$ is the anchoring strength of the cholesteric at the particle surface, and $K$ is the elastic constant of the liquid crystal (defined such that $W>0$ gives preferential perpendicular orientation of the director field at the particle surface). For the latter we choose $w = 0.23, 2.3$ which are within typical experimental range~\cite{tiffany}. This parameter can be viewed as the ratio of the anchoring energy, $WR^2$, to the elastic energy scale for distortion $KR$.

For a single colloidal particle in a nematic fluid, $w$ controls the formation of a local topological defect at the particle surface~\cite{stark}, with weak distortion of the nearby fluid at small $w$. In the colloid-BP system, $w$ also plays a determining role in the final composite structure. When $w$ is small (Fig.~ 1A,B), the attraction of particles to the disclination lattice causes
 the lattice to become covered by particles, with little disruption of its
 long-range order. Thus a templated colloidal crystal is created, as first reported in~\cite{miha}. Interestingly, the weak but finite anchoring apparently creates an elastic force between the colloids which leads to the formation of small colloidal lines within disclinations, especially for dilute samples (Fig. 1A). It seems likely that polymer-stabilised  blue phases~\cite{kikuchi} work by a similar principle to this small $w$ case, with a weak segregation of monomers towards the defect regions.

@@

A more practical protocol for including particles in a BPI phase is to disperse the particles in an isotropic phase, and then quenching this into the temperature range where BPI is stable. Here we know that, without particles, the kinetics can favour disordered (BPIII-like) intermediates which may be long lived or even metastable~\cite{domaingrowth}. Fig.~1E-H shows outcomes of such a quench protocol with the same parameter sets as in Fig.~1A-D. For small $w$, the particles again decorate the disclination lines, but are this time templated onto an amorphous network rather than an ordered one.
Increasing the volume fraction (Fig.~1F) leads to a more even coverage with
a better defined particle spacing. 
For large $w$, on the other hand, strong interparticle forces again dictate
the self-assembly. At low concentrations, discrete particle clusters decorate an amorphous defect lattice (Fig.~1G) whereas at higher $\phi$ a percolating network of colloid strands (the gel in Fig.~1H) is again seen. 
The structures in Fig. 1A-H have been followed for about 20 ms 
(see Supporting Information for the mapping between simulation and physical 
time); they correspond to deep metastable minima (elastic interactions are of
order $KR$ and dwarf thermal motion~\cite{stark,lavrentovich}).

In display devices and other optical applications, it is typical for liquid crystals to be strongly confined (and subject to strong boundary effects), in contrast to the bulk geometries considered above. We therefore next describe how colloidal suspensions self assemble in thin sandwiches (thickness $h = 3.5 \lambda$), whose confining walls can favour either tangential (planar degenerate) or normal (homeotropic) orientation of the director field. For simplicity we here impose one or the other as a strict boundary condition.

@@

An interesting question, relevant for applications to devices, is whether 
these structures can be modified, or switched, with electric fields.
Fig.~3 addresses this question, starting again from an amorphous blue phase,
this time at a higher chirality. Dispersing colloids in this networks
results in the particle occupying the available nodes, where the order
parameter is low (Fig.~3A). Applying a high enough field along the $z$ 
direction leads to a switching to a new phase, where colloidal arrays fit 
within a network with distorted hexagonal cells (Fig.~3B). Remarkably, upon 
field removal this honeycomb structure does not find its way back to the 
configuration before switching on the field, but gets stuck into a metastable 
structure (Fig. S7), with residual anisotropy along the field switching 
direction.
The resulting multistability is indeed a prerequisite for new generation 
switchable electrooptic devices for use in energy-saving electronics-free
appliances, smart glass, and e-paper to cite just a few examples.

@@

The surprising wealth of colloid-BP composites obtained in Figs.~1-3 is the complex consequence of a simple competition between two self-assembly principles. 
First, blue phases provide a three-dimensional template onto which colloidal particles are attracted so as to reduce the elastic stresses arising at the disclination cores~\cite{miha}. Second, colloidal intrusions in a liquid crystal raise the free energy locally; sharing this cost creates a generic tendency to aggregate into clusters~\cite{tiffany}. Which of these factors dominates depends on $w$, a dimensionless measure of the anchoring strength of the director field at the colloidal surface. For $w\ll 1$, the templating principle dominates. Colloidal crystals, gels or twisted rings can then arise. For $w \gg 1$, interparticle attractions defeat the templating tendency, and we predict disconnected aggregates, percolating gels, or helical colloidal ropes. In both cases, control is offered by the process route, colloid concentration, and the anchoring conditions at sample walls. We have also shown that electric fields can be used to switch between metastable states, providing a route to future multistable energy-saving devices. 
%This variety of metastable structures, with tunable properties that can combine the mechanical and optical features of colloids and liquid crystals~\cite{lavrentovich}, offers promise for the design of multistable liquid crystal devices, which are at the forefront of current technological developments in this field. 
We have not explored in this work their switchability by flow, but this could again be inherited from the underlying BP dynamics \cite{permeation2,adriano}, in which case they may also be relevant to the emerging technology of optofluidics~\cite{optofluidics}.

@@

\bibitem{tanakanatmat} T. Araki, M. Buscaglia, T. Bellini, H. Tanaka.
Memory and topological frustration in nematic liquid crystals confined in porous materials. {\it Nat. Mat.} {\bf 10}, 303-309 (2011).
\bibitem{fukuda} J.-I. Fukuda, S. Zumer.
Quasi-two-dimensional Skyrmion lattices in a chiral nematic liquid crystal. 
{\it Nat. Comm.} {\bf 2}, 246 (2011).

@@

%\bibitem{flowswitch} Davide et al on flow reorganization of BPs

@@

\begin{figure}
\centerline{\includegraphics[width=\textwidth]{text-fig3.jpg}}
\caption{Snapshots of the steady states obtained when a dispersion of colloids in the isotropic phase, then quenched into a regime where BPIII is stable in the bulk (A), and finally subjected to an electric field along the vertical direction (B). The electric field leads to the formation of a honeycomb structure with hexagonal ordering perpendicular to the field direction. 
Structures correspond to $w=0.23$ and $\phi=1\%$ (see Supporting Information
for full list of parameters). }
\end{figure}

>>>>

When spherical colloidal particles are mixed into a nematic liquid crystal, 
they disrupt the orientational order in the fluid and create defects (disclination lines) in the nematic close to their surfaces. To minimize the free energy cost, it is generally advantageous for defects to be shared:
therefore the colloidal particles have a generic tendency to aggregate. Such aggregates may further 
self-assemble into lines~\cite{wiresmiha}, 2D crystals~\cite{zumer}, 
planar structures~\cite{tanaka}, or 3D amorphous glasses~\cite{tiffany}.
Besides being of fundamental interest to materials science, these
structures have tunable elastic and optical properties~\cite{stark}. Hence 
they offer exciting prospects for applications as biosensors~\cite{abbott}, or
as new-generation devices~\cite{colloiddevice,electrophoresis}.

But one may also choose to disperse colloidal particles in a cholesteric, 
rather than nematic, liquid crystal. The molecules making up this material are
chiral, and this causes their average orientation (described by a director field) to rotate in space in
a helical fashion, rather than remain uniform as in nematics. 
This typically creates a 1-dimensional periodic structure, called the cholesteric phase, whose wavelength 
is the pitch, $p$. The periodicity can however become fully three dimensional in the so-called blue phases (BPs)~\cite{mermin}. BPs arise because the director field can twist around more than one direction at once (see 
the cartoon in Fig. S1). Such double twist 
regions (generally cylinders: DTCs) are energetically favoured at high enough molecular chirality, but
lead to geometric frustration because it is impossible to tile the whole 3D
space with DTCs without also introducing disclination lines: singular topological defects on which the director field is undefined. In blue phases these disclination lines themselves either form a 3D periodic regular lattice (in so called BPI and BPII), or remain disordered (BPIII)~\cite{bp3}. 
%
BPs are stabilised by the tendency of chiral molecules to twist, but destabilised by the cost of forming topological defects (which locally diminish the molecular alignment). In conventional cholesteric materials~\cite{mermin} this balance is only achievable in a narrow temperature range of a few degrees, around the onset of liquid crystalline ordering. However, recent advances in formulation have widened this stability range enormously (up to 250K) paving the way to
applications of BPs in display devices~\cite{kikuchi,bplasers,coleswidetrange,bpdevice,coles}.

@@

A second compelling reason to study colloid-BP composites is that, even without particles, BPs can show several competing metastable free energy minima,
corresponding to different topologies of the defect network~\cite{adriano}, with a strong dependence on applied electric or magnetic fields~\cite{henrichfield}.
% 
Adding particles is likely to further enrich the free energy landscape, creating composites that might be promising candidates for bistable or multistable devices, in which energy is needed only to switch optical properties and not to maintain them. (This is the e-paper paradigm~\cite{epaper}, and can lead to huge energy savings.)
To explore such metastability, and switching strategies between states, we require multi-unit-cell, time-dependent simulations. These go beyond previous studies of colloid-BP composites~\cite{miha} but follow comparable simulations of pure BPs~\cite{bp3,henrichfield,domaingrowth}, and isolated and dimeric colloids in cholesterics~\cite{juho1,juho2}. Our methodology and simulation parameters are summarized in the Supporting Information.

We show in Fig.~1 A-D the final state structures of four simulations in which we first equilibrated a stable BPI disclination network in bulk (with periodic boundary conditions) and then dispersed colloidal particles within it. Among dimensionless control parameters are the colloid volume fraction $\phi$  ($\phi = 1\%$ or $5\%$); the ratio $r = R/\lambda$ of particle radius to the BP lattice parameter $\lambda$ (here $r\simeq 0.14$) and the ratio $w = WR/K$ where $W$ is the anchoring strength of the cholesteric at the particle surface, and $K$ is the elastic constant of the liquid crystal (defined such that $W>0$ gives preferential perpendicular orientation of the director field at the particle surface). For the latter we choose $w = 0.23, 2.3$ which are within typical experimental range~\cite{tiffany}. This parameter can be viewed as the ratio of the anchoring energy, $WR^2$, to the elastic energy scale for distortion $KR$.

For a single colloidal particle in a nematic fluid, $w$ controls the formation of a local topological defect at the particle surface~\cite{stark}, with weak distortion of the nearby fluid at small $w$. In the colloid-BP system, $w$ also plays a determining role in the final composite structure. When $w$ is small (Fig.~ 1A,B), the attraction of particles to the disclination lattice causes
 the lattice to become covered by particles, with little disruption of its
 long-range order. Thus a templated colloidal crystal is created, as first reported in~\cite{miha}. Interestingly, the weak but finite anchoring apparently creates an elastic force between the colloids which leads to very slow dynamics (Fig. S4) and the formation of small colloidal lines within disclinations, especially for dilute samples (Fig. 1A). It seems likely that polymer-stabilised  blue phases~\cite{kikuchi} work by a similar principle to this small $w$ case, with a weak segregation of monomers towards the defect regions.

@@

A more practical protocol for including particles in a BPI phase is to disperse the particles in an isotropic phase, and then quenching this into the temperature range where BPI is stable. Here we know that, without particles, the kinetics can favour disordered (BPIII-like) intermediates which may be long lived or even metastable~\cite{domaingrowth}. Fig.~1E-H shows outcomes of such a quench protocol with the same parameter sets as in Fig.~1A-D. For small $w$, the particles again decorate the disclination lines, but are this time templated onto an amorphous network rather than an ordered one.
Increasing the volume fraction (Fig.~1F) leads to a more even coverage with
a better defined particle spacing. 
For large $w$, on the other hand, strong interparticle forces again dictate
the self-assembly. At low concentrations, discrete particle clusters decorate an amorphous defect lattice (Fig.~1G) whereas at higher $\phi$ a percolating network of colloid strands (the gel in Fig.~1H) is again seen. 
The structures in Fig. 1A-H have been followed for $\sim$20 ms 
(see Fig. S4); they correspond to either very slowly evolving states or 
deep metastable minima (elastic interactions are of
order $KR$ and dwarf thermal motion, see
Fig. S4 and Ref.~\cite{stark,lavrentovich}).

In display devices and other optical applications, it is typical for liquid crystals to be strongly confined (and subject to strong boundary effects), in contrast to the bulk geometries considered above. We therefore next describe how colloidal suspensions self assemble in thin sandwiches (thickness $h = 1.75 \lambda$), whose confining walls can favour either tangential (planar degenerate) or normal (homeotropic) orientation of the director field. For simplicity we here impose one or the other as a strict boundary condition.

@@

An interesting question, relevant for applications to devices, is whether 
these structures can be modified, or switched, with electric fields.
Fig.~3 addresses this question, starting from amorphous Blue Phase III,
this time at a higher chirality. Dispersing colloids in this networks
results in the particle occupying the available nodes, where the order
parameter is low (Fig.~3A). Applying a high enough field along the $z$ 
direction leads to a switching to a new phase, where colloidal arrays fit 
within a network with distorted hexagonal cells (Fig.~3B). Remarkably, upon 
field removal this honeycomb structure does not find its way back to the 
configuration before switching on the field, but gets stuck into a metastable 
structure (Fig. S7), with residual anisotropy along the field switching 
direction.
The resulting multistability is indeed a prerequisite for new generation 
switchable electrooptic devices for use in energy-saving electronics-free
appliances, smart glass, and e-paper to cite just a few examples.

@@

The surprising wealth of colloid-BP composites obtained in Figs.~1-3 is the complex consequence of a simple competition between two self-assembly principles. 
First, blue phases provide a three-dimensional template onto which colloidal particles are attracted so as to reduce the elastic stresses arising at the disclination cores~\cite{miha}. Second, colloidal intrusions in a liquid crystal raise the free energy locally; sharing this cost creates a generic tendency to aggregate into clusters~\cite{tiffany}. Which of these factors dominates depends on $w$, a dimensionless measure of the anchoring strength of the director field at the colloidal surface. For $w\ll 1$, the templating principle dominates. Colloidal crystals, gels or twisted rings can then arise. For $w \gg 1$, interparticle attractions defeat the templating tendency, and we predict disconnected aggregates, percolating gels, or helical colloidal ropes. In both cases, control is offered by the process route, colloid concentration, and the anchoring conditions at sample walls. We have also shown that electric fields can be used to switch between metastable states, providing a route to future multistable energy-saving devices. 
%This variety of metastable structures, with tunable properties that can combine the mechanical and optical features of colloids and liquid crystals~\cite{lavrentovich}, offers promise for the design of multistable liquid crystal devices, which are at the forefront of current technological developments in this field. 
We have not explored in this work their switchability by flow \cite{flowswitch}, but this could again be inherited from the underlying BP dynamics \cite{adriano,permeation2}, in which case they may also be relevant to the emerging technology of optofluidics~\cite{optofluidics}.

@@

%\bibitem{tanakanatmat} T. Araki, M. Buscaglia, T. Bellini, H. Tanaka.
%Memory and topological frustration in nematic liquid crystals confined in porous materials. {\it Nat. Mat.} {\bf 10}, 303-309 (2011).
%\bibitem{fukuda} J.-I. Fukuda, S. Zumer.
%Quasi-two-dimensional Skyrmion lattices in a chiral nematic liquid crystal. 
%{\it Nat. Comm.} {\bf 2}, 246 (2011).

@@

\bibitem{flowswitch} O. Henrich, K. Stratford, P.V. Coveney, M.E. Cates, D. Marenduzzo,
Rheology of cubic blue phases. {\it Soft Matter} (2013), DOI: 10.1039/C3SM50228G 

@@

\begin{figure}
\centerline{\includegraphics[width=\textwidth]{text-fig3.jpg}}
\caption{Snapshots of the steady states obtained when a dispersion of colloids is inserted into a thermodynamically stable bulk-BPIII (A), and finally subjected to an electric field along the vertical direction (B). The electric field leads to the formation of a honeycomb structure with hexagonal ordering perpendicular to the field direction. 
Structures correspond to $w=0.23$ and $\phi=1\%$ (see Supporting Information
for full list of parameters). }
\end{figure}
